UKB Simulation Analyses

### 2023-12-05 ukb_sim.site_analyses.Rmd
reorg

## Raw Data Results
Review outcomes from raw (non-ComBat-ed) data

```{r, fig.width=11}
sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & dataset ==	"ukb-CN-data-simsites") %>%
  ggplot() +
  geom_col(aes(x=reorder(pheno,t_stat), y=t_stat, fill=pheno_cat)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & dataset == "ukb-CN-data-simsites") %>%
  ggplot() +
  geom_col(aes(x=label, y=t_stat, fill=pheno_cat)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & dataset == "ukb-CN-data-simsites") %>%
  ggplot() +
  geom_col(aes(x=region, y=t_stat, fill=hemi)) + 
  facet_wrap(~pheno_cat) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Significant effects of male sex on sigma across phenotype-specific brain charts")
```

Plot on brains!

```{r}
sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & dataset == "ukb-CN-data-simsites") %>%
  group_by(pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_wrap(~pheno_cat) +
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")
```

## ComBat Effects by Moment

@@

@@

var.long %>%
  dplyr::filter(!(feature %in% vol_list_global) & (Source_File == "cf.gam_data.csv" |Source_File == "ukb_CN_data_simsites.csv" |  Source_File == "cf.gamlss_data.csv" )) %>%
  ggplot() +
  geom_density(aes(x=variance, fill=Source_File), alpha = 0.5, position="identity") +
  facet_wrap(~sim.site)

#Source_File == "cf.lm_data.csv" | 
```

@@

Simple t test for sex effects across combat types
```{r}
sex_ttest_df <- data.frame("Source_File" = levels(combined_data$Source_File))

@@

Basically same results gam/gamlss

@@

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")
```
Check number of regions w significant sex effects in sigma
```{r}
sex.sigma.counts <- sex.df %>%
  dplyr::filter(term=="sexMale" & parameter == "sigma") %>%
  group_by(dataset, pheno_cat) %>%
  summarize(n_sig_regions = sum(sig.drop_bf.corr),
            n_regions = n())
sex.sigma.counts
```

More variation
```{r}
ggplot(sex.sigma.counts) +
  geom_col(aes(x=dataset, y=n_sig_regions, fill=dataset)) +
  facet_wrap(~pheno_cat) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

##### Variance
Convert sigma to variance by correcting for log-link function (sex's effect on variance = exp(sigma) - exp(log(sigma) - beta sex)) -> exp(beta_sex).
Also getting mean variance.

@@

sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_histogram(aes(x=var_est, fill = dataset), color="black", alpha=0.5, position = "identity") +
  theme_linedraw() +
  facet_wrap(~pheno_cat, scales= "free")
```

```{r, fig.height=10, fig.width=8}
proportion_beta_plot <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Proportion of effect of male sex on variance")
```

@@

ggsave("male_variance_beta.jpeg", beta.plot, width=12, height=15)
```

@@

fix age, get diff between male and female pred. variance
```{r}
pred.csvs <- list.files(path = "ukb_basic", pattern = "_variance.csv", full.names = TRUE)
combined_var <- data.frame()

@@

table(combined_var$dataset)
#skim(combined_preds)
```

```{r}
combined_var <- combined_var %>%
  mutate(sex_std = sex_effect/f.var)

summary(sigma.df$var_est)
summary(combined_var$sex_effect)

summary(sigma.df$beta_ratio)
summary(combined_var$sex_std)
```

```{r}
plot.info <- sigma.df %>%
  dplyr::filter(term == "sexMale") %>%
  dplyr::select(mod_name, pheno, dataset, pheno_cat, label, hemi, region, sig.sum_bf.corr) %>%
  unique() %>%
  mutate(dataset = gsub("-", "_", dataset),
         dataset = gsub("_simsites", "", dataset))
combined_var <- inner_join(combined_var, plot.info)
table(combined_var$dataset)
```

@@

```{r, fig.height=5, fig.widt=5}
combined_var %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = sex_std)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Standardized effect of male sex on variance")
```

using aseg

```{r}
plot(aseg)
```

## ComBat Effects on Centiles

@@

table(combined_preds$dataset)
#skim(combined_preds)
```
calc how much centiles change
```{r}
cent_diffs <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "abs.diff_{.col}")) %>%
  ungroup()

diff_vol_glob <- sapply(vol_list_global, function(x) paste0("abs.diff_", x))

mean.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup()

@@

Plot average differences in centiles
```{r}
mean.diffs.long <- mean.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
mean.diffs.long <- left_join(mean.diffs.long, dk.parc, by="label")
```

@@

Look at max difference per phenotype
```{r, fig.height=6, fig.width=7}
max.diffs.long <- max.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

Test for differences in distribution
```{r}
#get centile errors for each subject
cent.diffs2 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref

@@

@@

  df.pairwise.t <- tidy(pairwise.t.test(x=cent.diffs2[[pheno]], g=cent.diffs2[["dataset"]], paired = TRUE, p.adj = "none"))
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()
```

@@

## ComBat Effects on Z-Score
calc how much z-scores change

@@

```{r}
beta.plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = var_est)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Oslo") +
  ggtitle("Beta weight for effect of male sex on variance")

>>>>

## By Moment

@@

Plot distribution of features' variances across each site

@@

var.long %>%
  dplyr::filter(!(feature %in% vol_list_global) & (Source_File == "cf.gam_data.csv" |Source_File == "ukb_CN_data_simsites.csv" |  Source_File == "cf.gamlss_data.csv" )) %>%
  ggplot() +
  geom_density(aes(x=variance, fill=Source_File), alpha = 0.5, position="identity") +
  facet_wrap(~sim.site)
```
How does male vs. female variance across features change by combat type

@@

Simple t test for sex effects on brain features to see whether # of significantly sex-biased features change across combat configurations

```{r}
sex_ttest_df <- data.frame("Source_File" = levels(combined_data$Source_File))

@@

Basically no difference across combat types

@@

sex.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & parameter == "sigma" & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = t_stat)) +
  scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_void() +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank()) +
  ggtitle("t-statistic for effect of male sex on sigma")
```
Check number of regions w significant sex effects in sigma

```{r}
sex.sigma.counts <- sex.df %>%
  dplyr::filter(term=="sexMale" & parameter == "sigma") %>%
  group_by(dataset, pheno_cat) %>%
  summarize(n_sig_regions = sum(sig.drop_bf.corr),
            n_regions = n())
sex.sigma.counts
```
 Plot these results

```{r}
ggplot(sex.sigma.counts) +
  geom_col(aes(x=dataset, y=n_sig_regions, fill=dataset)) +
  facet_wrap(~pheno_cat) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("# regions with significant effects of male sex on sigma")
```

## Variance
Estimating by 3 different methods:
1. Convert sigma to variance by correcting for log-link function (sex's effect on variance = exp(sigma) - exp(log(sigma) - beta sex)) -> exp(beta_sex).
1.5. getting mean variance exp(mean sigma) (not actually plotted)
2. actual estimated variance at mean(sample age) (andrew's rec)

#### Method 1: exp(sigma)

@@

sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_histogram(aes(x=var_est, fill = dataset), color="black", alpha=0.5, position = "identity") +
  theme_linedraw() +
  facet_wrap(~pheno_cat, scales= "free")+
  ggtitle("estimated variance in regions with significant effects of male sex on sigma")
```

What proportion of estimated variance is attributable to sex?

```{r, fig.height=10, fig.width=8}
proportion_beta_plot <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = beta_ratio)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=15) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork") +
  ggtitle("Proportion of effect of male sex on variance")
```

exp(SexMale beta weight on sigma) -> estimate of male effect on variance

@@

ggsave("figs/male_variance_beta.jpeg", beta.plot, width=12, height=15)
```

@@

#### Method 2: Var at Mean Age
fix age, get diff between male and female pred. variance
```{r}
pred.csvs <- list.files(path = "ukb_basic", pattern = "_variance.csv", full.names = TRUE)
combined_var <- data.frame()

@@

table(combined_var$dataset) #confirm all data's there
#skim(combined_preds)
```

Calculate standardized effects of male sex w/in region as (M var - F var)/F var

```{r}
combined_var <- combined_var %>%
  mutate(sex_std = sex_effect/f.var) #standardized effect of male sex on variance

#compare methods
summary(sigma.df$var_est) #method 1 var
summary(combined_var$sex_effect) #method 2 var

summary(sigma.df$beta_ratio). #method 1 male effect on var
summary(combined_var$sex_std) #method 2 male effect on var
```

```{r}
#join with sigma.df for significance testing
plot.info <- sigma.df %>%
  dplyr::filter(term == "sexMale") %>%
  dplyr::select(mod_name, pheno, dataset, pheno_cat, label, hemi, region, sig.sum_bf.corr) %>%
  unique() %>%
  mutate(dataset = gsub("-", "_", dataset),
         dataset = gsub("_simsites", "", dataset))
combined_var <- inner_join(combined_var, plot.info)
table(combined_var$dataset) #check no data's lost
```

@@

```{r, fig.height=5, fig.widt=5}
combined_var %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume") %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = sex_std)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=12) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork", name = "Sex Effect", limits = c(-0.12, 0.12), breaks = c(-0.1, -0.05, 0, 0.05, 0.1)) +
  ggtitle("Standardized effect of male sex on variance")
```

## Centiles

@@

table(combined_preds$dataset)
#skim(combined_preds)
```

#### Magnitude error

Calculate how far centiles are from "truth" (raw data's centile vals), take absolute value for magnitude error

```{r}
cent_diffs <- combined_preds %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "abs.diff_{.col}")) %>%
  ungroup()
```

Calculate within feature mean and max abs. centile errors across all subjects

```{r}
mean.cent.diffs <- cent_diffs %>%
  group_by(dataset) %>%
  summarize(across(starts_with("abs.diff_"), mean)) %>%
  ungroup()

@@

Plot w/in feature average magnitude of centile errors
```{r}
mean.diffs.long <- mean.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)),
      label = sub("_[^_]*_", "_", pheno))
mean.diffs.long <- left_join(mean.diffs.long, dk.parc, by="label")
```

@@

Look at max difference per phenotype

```{r, fig.height=6, fig.width=7}
max.diffs.long <- max.cent.diffs %>%
  pivot_longer(cols=starts_with("abs.diff_"), values_to="max_centile_abs.diff", names_to = "pheno", names_prefix = "abs.diff_") %>%
  mutate(pheno_cat = as.factor(case_when(
      pheno %in% vol_list_global ~ "Global Volume",
      pheno %in% vol_list_regions ~ "Regional Volume",
      pheno %in% sa_list ~ "Regional SA",
      pheno %in% ct_list ~ "Regional CT",
      TRUE ~ NA_character_)))

@@

Test for differences in distribution of subjects' abs(cent error)

```{r}
#get abs val of centile errors for each subject
cent.diffs2 <- combined_preds %>%
  dplyr::filter(!str_detect(dataset, "_refA")) %>%
  mutate(dataset = factor(dataset, levels = unique(dataset), ordered = FALSE)) %>%
  mutate(dataset = relevel(dataset, ref= "ukb_CN")) %>%
  arrange(dataset) %>%
  group_by(participant) %>%
  mutate(across(all_of(pheno_list), ~ abs(. - first(.)), .names = "diff_{.col}")) %>%
  ungroup() %>%
  dplyr::filter(dataset != "ukb_CN") #drop ref

@@

plot to see if assumptions for t-tests hold ...
```{r}
ggplot(cent.diffs2) +
  geom_histogram(aes(x=diff_rh_CT_posteriorcingulate, fill=dataset), alpha = 0.6)
```

looks like.... no.

@@

  df.pairwise.t <- tidy(pairwise.wilcox.test(x=cent.diffs2[[pheno]], g=cent.diffs2[["dataset"]], paired = TRUE, p.adj = "none")) #update to wilcox since t assumptions not all met
  df.pairwise.t$pheno <- pheno
  t.df <- rbind(t.df, df.pairwise.t)
}
detach()
```

@@

## Z-Score
calc how much z-scores change

@@

```{r}
beta.plot2 <- sigma.df %>%
  dplyr::filter(sig.sum_bf.corr == TRUE & pheno_cat != "Global Volume" & term =="sexMale") %>%
  filter(!str_detect(dataset, "-refA")) %>%
  mutate(dataset = factor(dataset, levels = c("ukb-CN-simsites", "cf", "cf.lm", "cf.gam", "cf.gamlss"))) %>%
  group_by(dataset, pheno_cat) %>%
  ggplot() +
  geom_brain(atlas = dk, 
             position = position_brain(hemi ~ side),
             aes(fill = var_est)) +
  #scale_fill_gradient2() +
  facet_grid(dataset~pheno_cat) + 
  theme_bw(base_size=20) +
  theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.border = element_rect(color= "gray", fill = NA)) +
  paletteer::scale_fill_paletteer_c("grDevices::Cork") +
  ggtitle("Beta weight for effect of male sex on variance")
